---
title: "Hypertension Risk Atlas: Data Ingestion- CDC Places"
author: "Nancy Walker"
format: 
  pdf:
    keep-tex: true
    include-in-header:
      text: |
        \usepackage{float}      % enables [H]
        \usepackage{placeins}   % provides \FloatBarrier
        \floatplacement{figure}{H}
        \floatplacement{table}{H}
bibliography: references.bib
csl: https://raw.githubusercontent.com/citation-style-language/styles/master/apa.csl
execute:
  echo: true
  results: asis                    # show results inline
  warning: false
  message: false

---

# Data Validation: CDC PLACES Prevalence Metrics

Initial validation of the CDC PLACES dataset revealed that the raw API stream provides exclusively Crude prevalence metrics for all health measures, including our target variable: High blood pressure among adults.

# Data Limitation & Strategic Context

This discovery introduces a necessary limitation in our current analysis. Because Crude prevalence is not adjusted for age, our current dataset reflects the raw, unadjusted health burden of a population as it exists today. Consequently, counties with older demographic profiles will naturally exhibit higher reported prevalence rates for hypertension.

To better clarify the distinction:

    Crude Prevalence: Highly valuable for immediate resource allocation. Policymakers use this to identify where the highest burden of disease exists, regardless of the underlying demographic drivers.

    Age-Adjusted Prevalence: Better suited for health disparity and inequality research. By neutralizing age as a variable, this metric allows for more accurate comparisons between counties with vastly different demographic structures, making it easier to isolate systemic socioeconomic or environmental drivers of health.

# Next Steps for the Machine Learning Pipeline

Our project goal is to build a spatial machine-learning pipeline that predicts county-level hypertension prevalence using social and environmental determinants. Our primary objective is to mitigate the risk of resource misallocation and address widening health disparities.

Given this objective, retrieving the Age-Adjusted prevalence data is a critical next step. While the current Crude dataset serves as a useful baseline for exploratory analysis, we will transition to the Age-Adjusted dataset in the upcoming iteration to ensure our predictive model identifies "excess risk"—areas where hypertension prevalence is high even after controlling for age—thereby isolating the true impact of the social determinants we are investigating.

# Import Libraries 

```{r}
library(RSocrata)   # For CDC PLACES
library(tidyverse)
```

## DIRECTORY MANAGEMENT

```{r}
# Define Data Folder Stucture
data_folders <- c("data/raw", "data/processed", "data/final")

# Loop through and create only if they don't exist
for (folder in data_folders) {
  if (!dir.exists(folder)) {
    dir.create(folder, recursive = TRUE)
    message(paste0("📁 Created folder: ", folder))
  } else {
    message(paste0("📂 Folder already exists: ", folder))
  }
}

```

#  DATA DOWNLOAD / INGESTION

## CDC PLACES Data 

```{r}
# Define file path first
local_cdc_aa_csv <- "data/raw/cdc_places_aa_raw.csv"

# Define your base endpoint (Example ID: swc5-untb for County data)
base_url <- "https://data.cdc.gov/resource/swc5-untb.csv"

# Construct the filter string
# Note: Use backticks for column names and ensure the values match exactly
filter_query <- "?$where=`data_value_type`='Age-adjusted prevalence'"

# Combine into your final endpoint
cdc_endpoint <- paste0(base_url, filter_query)

# Now your defensive check remains the same
if (file.exists(local_cdc_aa_csv)) {
  message(" Local CDC PLACES cache found! Loading data directly from disk...")
  cdc_places_raw <- readr::read_csv(local_cdc_aa_csv, show_col_types = FALSE)
  
} else {
  message("Connecting to CDC PLACES API with filter... Fetching subset...")
  # read.socrata handles the combined URL automatically
  cdc_places_aa_raw <- read.socrata(cdc_endpoint)
  
  readr::write_csv(cdc_places_aa_raw, local_cdc_aa_csv)
  message("Filtered CDC PLACES dataset saved.")
}
```

```{r}
# Extract a small subset to verify data parity in the report preview
cdc_test <- cdc_places_aa_raw %>%
  head(5) %>%
  dplyr::select(1:5)

# Render a clean preview table for the PDF report
knitr::kable(cdc_test, caption = "CDC PLACES Sample Data Stream (First 5 Columns)")
```

