---
title: "Hypertension Risk Atlas: Data Ingestion"
author: "Nancy Walker"
format: 
  pdf:
    keep-tex: true
    include-in-header:
      text: |
        \usepackage{float}      % enables [H]
        \usepackage{placeins}   % provides \FloatBarrier
        \floatplacement{figure}{H}
        \floatplacement{table}{H}
bibliography: references.bib
csl: https://raw.githubusercontent.com/citation-style-language/styles/master/apa.csl
execute:
  echo: true
  results: asis                    # show results inline
  warning: false
  message: false

---

# R Data Ingestion 

Build automated R scripts to programmatically fetch and ingest four core datasets: the **USDA Food Access Research Atlas**, the **CDC PLACES** county-level health indicators, the **County Health Rankings & Roadmaps (CHR&R)** metrics, and the **U.S. Census Bureau ACS** socioeconomic data.

## 1. SETUP & LIBRARIES

A few specialized libraries will be needed to fetch the data from the various sources. 

```{r}
# Install any missing standard CRAN packages if needed
list.of.packages <- c("readxl", "tidyverse", "tidycensus", "kableExtra", "stringr", "dplyr", "tidyr")
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]

if(length(new.packages)) {
  install.packages(new.packages, repos = "https://cloud.r-project.org")
}

# Handle archived RSocrata via GitHub (Only runs if missing)
if (!requireNamespace("RSocrata", quietly = TRUE)) {
  if (!requireNamespace("remotes", quietly = TRUE)) {
    install.packages("remotes", repos = "https://cloud.r-project.org")
  }
  remotes::install_github("Chicago/RSocrata", upgrade = "never") 
}
```

```{r}
library(RSocrata)   # For CDC PLACES
library(readxl)     # For reading/verifying USDA Excel files
library(tidyverse)  # For data writing utilities (readr)
library(tidycensus) # For Censue Bureau data 
library(knitr)
library(kableExtra)
library(stringr)
library(dplyr)
library(tidyr)
```

## 2. DIRECTORY MANAGEMENT

```{r}
# Define Data Folder Stucture
data_folders <- c("data/raw", "data/processed", "data/final")

# Loop through and create only if they don't exist
for (folder in data_folders) {
  if (!dir.exists(folder)) {
    dir.create(folder, recursive = TRUE)
    message(paste0("📁 Created folder: ", folder))
  } else {
    message(paste0("📂 Folder already exists: ", folder))
  }
}

```

## 3. DATA DOWNLOAD / INGESTION

### CDC PLACES Data 

Ingest county-level chronic disease estimates from the CDC PLACES dataset [@cdc_places_2025].

```{r}
# CDC PLACES County-level Data API Endpoint (Socrata Open Data API)
# This points directly to the latest release stream
cdc_endpoint <- "https://data.cdc.gov/resource/cwsq-ngmh.csv"
local_cdc_csv <- "data/raw/cdc_places_raw.csv"
```

```{r}
# Defensive Check: Skip API network call if the raw file is already cached locally
if (file.exists(local_cdc_csv)) {
  message(" Local CDC PLACES cache found! Loading data directly from disk...")
  cdc_places_raw <- readr::read_csv(local_cdc_csv, show_col_types = FALSE)
  
} else {
  message("Connecting to CDC PLACES API... Fetching full dataset (this may take 1-3 minutes)...")
  cdc_places_raw <- read.socrata(cdc_endpoint)
  
  # Persist Raw Stream: Immediately write the unmodified data to local storage
  readr::write_csv(cdc_places_raw, local_cdc_csv)
  message("Full CDC PLACES dataset successfully saved to data/raw/cdc_places_raw.csv")
}
```

```{r}
# Extract a small subset to verify data parity in the report preview
cdc_test <- cdc_places_raw %>%
  head(5) %>%
  dplyr::select(1:5)

# Render a clean preview table for the PDF report
knitr::kable(cdc_test, caption = "CDC PLACES Sample Data Stream (First 5 Columns)")
```

### SUSDA Food Access Research Atlas

The USDA server restricts automated programmatic downloads of census-tract-level food accessibility metrics from the USDA Economic Research Service. This dataset must be placed in the raw data directory manually [@usda_atlas_2019].

```{r}
# Define the exact path where the file must live
local_xlsx <- "data/raw/FoodAccessResearchAtlasData2019.xlsx"

# Defensive Check: If the file is missing, halt and print clean instructions
if (!file.exists(local_xlsx)) {
  stop(paste0(
    "\n❌ MISSING REQUIRED DATASET!\n",
    "Please follow these steps to add the USDA Food Atlas data:\n",
    "1. Go to: https://www.ers.usda.gov/data-products/food-access-research-atlas/download-the-data#:~:text=Food%20Access%20Research%20Atlas%20Data%20Download%202019 \n",
    "2. Click 'Download Food Access Research Atlas Data Download 2019 (XLSX, 81.83 MB)' under the Current Version header.\n",
    "3. Move the downloaded file into this project folder under: data/raw/\n",
    "4. Rename it exactly to: FoodAccessResearchAtlasData2019.xlsx\n\n"
  ))
}

# If the file exists, safely read it into the active R session
message("✅ USDA dataset found! Ingesting data...")
usda_raw <- readxl::read_excel(local_xlsx, sheet = "Food Access Research Atlas")

# Extract a small subset to verify data parity
usda_test <- usda_raw %>%
  head(5) %>%
  dplyr::select(State, County, CensusTract, LowIncomeTracts, LAPOP1_10)

# Render a clean preview table
knitr::kable(usda_test, caption = "USDA Food Access Research Atlas Sample Data")
```

### County Health Rankings & Roadmaps (CHR&R)

The CHR&R repository provides comprehensive county-level health determinants. This dataset must be placed in the raw data directory manually [@chrr_2025].

```{r}
# Define the exact path where the file must live
chrr_csv <- "data/raw/analytic_data2025.csv"

# Defensive Check: If the file is missing, halt and print instructions
if (!file.exists(chrr_csv)) {
  stop(paste0(
    "\n\n❌ MISSING REQUIRED DATASET!\n",
    "Please follow these steps to add the CHR&R data:\n",
    "1. Go to: https://www.countyhealthrankings.org/health-data/methodology-and-sources/data-documentation \n",
    "2. Under the '2025 Annual Data Release' section, download the '2025 CHR CSV Analytic Data'.\n",
    "3. Move the file into this project folder under: data/raw/\n",
    "4. Rename it exactly to: analytic_data2025.csv\n\n"
  ))
}

# If the file exists, safely read it into the active R session
message("✅ CHR&R dataset found! Ingesting data...")
chrr_raw <- readr::read_csv(chrr_csv, skip = 1, show_col_types = FALSE)

# Extract a small subset to verify data parity
chrr_test <- chrr_raw %>%
  head(5) %>%
  dplyr::select(state, county, v001_rawvalue)

# Render a clean preview table
knitr::kable(chrr_test, caption = "CHR&R Health Policy Registry Sample Data")
```

### Census data

To ingest census data you will need an API key from the Census Bureau [@census_acs]
Get one here: https://api.census.gov/data/key_signup.html

In your R console, run the following command to set your key (replace "YOUR_KEY_HERE" with your actual key):
# Replace 'YOUR_KEY_HERE' with the key they emailed you
census_api_key("YOUR_KEY_HERE", install = TRUE, overwrite = TRUE)


```{r}
### Ingest U.S. Census Bureau (ACS 5-Year Estimates)

# Define variables we want (e.g., Median Income and Poverty Rate)
census_vars <- c(
  med_income = "B19013_001",
  poverty_rate = "B17001_002"
)

# Fetch data at the County level
message("Fetching Census ACS data...")
census_raw <- get_acs(
  geography = "county",
  variables = census_vars,
  year = 2023,
  survey = "acs5"
)

# Clean and pivot to a "wide" format (One row per FIPS)
census_clean <- census_raw %>%
  select(GEOID, variable, estimate) %>%
  pivot_wider(names_from = variable, values_from = estimate) %>%
  rename(fips = GEOID, 
         median_income = med_income, 
         poverty_count = poverty_rate)

```

```{r}
# Render a clean preview table of the Census data
census_clean %>%
  head(10) %>% # Shows only the first 10 rows
  kable(caption = "Census ACS 5-Year Estimates (Sample Data)") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

# Save the raw, unclearned API response into local storage for reproducibility
write_csv(census_clean, "data/raw/census_acs_county_2023.csv")
message("✅ Census ACS data saved to data/raw/census_acs_county_2023.csv")
```

## 5.DATA CLEANING & STANDARDIZATION

Ensure your 5-digit FIPS codes are treated as strings (characters), not numbers! 
(Hint: Numerical rendering drops leading zeros, turning "01001" into 1001, which will break your spatial match). 
Look into stringr::str_pad() if needed.

```{r}
# Create a summary of dataset dimensions
data_summary <- tibble::tibble(
  Dataset = c("CDC PLACES", "Census ACS", "CHR&R", "USDA Food Access"),
  Rows = c(nrow(cdc_places_raw), nrow(census_raw), nrow(chrr_raw), nrow(usda_raw)),
  Columns = c(ncol(cdc_places_raw), ncol(census_raw), ncol(chrr_raw), ncol(usda_raw)) 
)

# Render as a beautiful table for your report
knitr::kable(data_summary, caption = "Data Pipeline: Dataset Dimensions")
```

CDC PLACES is the largest file wutg over 3 million rows. It is likely broken into specific health indicators or subpopulations. For a county level model, this dartaset may need to be filtered to a specific heath indicator (e.g., hypertension) or a specific subpopulation (e.g., adults aged 18+). 

CHAR&R is 3,204 rows, which is perfect for our purposes since there are 3,142 counties in the U.S. (and some extra rows for territories). Country-level granularity is effectively captured. 

USDA Food Access is at census tract level, which is more granular than county. We will need to aggregate this data up to the county level before merging.

Census ACS had 6,444 rows. This is interesting becuse it is double the number of counties in the U.S. (3,142). This is because the ACS data is broken into two rows per county, or there is two years of data. We will need to filter this data to the most recent year and aggregate it to county level before merging.

```{r}
# Diagnostic: Verify FIPS columns before merging 
check_fips <- function(df, col_name) {
  # Check if FIPS column exists and is character type
  if (col_name %in% names(df)) {
    is_char <- is.character(df[[col_name]])
    has_5_digits <- all(nchar(as.character(df[[col_name]])) == 5)
    
    message(paste0("🔍 Checking ", col_name, ": Is Character? ", is_char, " | All 5 Digits? ", has_5_digits))
  } else {
    message(paste0("❌ Column '", col_name, "' not found!"))
  }
}

# Run this for each dataset
check_fips(cdc_places_raw, "StateCountyFIPS") # Update name to match your specific column
check_fips(usda_raw, "FIPS")
check_fips(chrr_raw, "fipscode")
check_fips(census_raw, "GEOID")
```

census_raw and chrr_raw are confirmed to be perfectly formatted.

Correct names for cdc_places_raw and usda_raw may need to be updated based on the actual column names in those datasets.

```{r}
# Reveal the hidden column names for CDC PLACES and USDA datasets
colnames(cdc_places_raw)
colnames(usda_raw)
```

```{r}
# Update names in the check_fips function calls based on the actual column names revealed above.

# CDC check
check_fips(cdc_places_raw, "countyfips") 

# USDA check (Note: This will likely fail or show length 11, which confirms it's a Tract)
check_fips(usda_raw, "CensusTract")
```

countyfips (CDC): Is currently numeric (Is Character? FALSE), which is why it lost its leading zeros (explaining why "All 5 Digits" is FALSE).

CensusTract (USDA): Is character, but its length is longer than 5 (which is expected for tracts).

Need to normalize them to a consistent 5-character string format for a perfect merge.

```{r}
# Standardize CDC PLACES
cdc_clean <- cdc_places_raw %>%
  mutate(GEOID = str_pad(as.character(countyfips), 5, side = "left", pad = "0"))

# Standardized USDA (Aggregated to county level)
usda_clean <- usda_raw %>%
  mutate(GEOID = str_sub(as.character(CensusTract), 1, 5)) %>%
  group_by(GEOID) %>%
  summarise(across(where(is.numeric), \(x) mean(x, na.rm = TRUE)), .groups = "drop")

# Standardize CHR&R
chrr_clean <- chrr_raw %>%
  mutate(GEOID = str_pad(as.character(fipscode), 5, side = "left", pad = "0"))

# Standardize Census ACS (Already using GEOID)
census_clean <- census_raw %>%
  mutate(GEOID = str_pad(as.character(GEOID), 5, side = "left", pad = "0"))
```

```{r}
# One final diagnostic to prove it works 
# This verifies all 4 are ready to merge
check_fips(cdc_clean, "GEOID")
check_fips(usda_clean, "GEOID")
check_fips(chrr_clean, "GEOID")
check_fips(census_clean, "GEOID")
```

All four datasets share the same GEOID key and can be combined into one master_df.

## 6. MERGING & EXPORT
### Join tables by FIPS key and write the clean output to a "data/processed/" directory.

```{r}
# Prepare Census: Pivot wide so each county is one row
census_wide <- census_clean %>%
  select(GEOID, variable, estimate) %>%
  pivot_wider(names_from = variable, values_from = estimate)

# Prepare CDC: Pivot wide so one row per county
# Note: Ensure 'measure' column contains unique values for each row
cdc_wide <- cdc_clean %>%
  group_by(GEOID, measure) %>%
  summarise(data_value = mean(data_value, na.rm = TRUE), .groups = "drop") %>%
  
  # Pivot to wide format safely
  pivot_wider(names_from = measure, values_from = data_value)

# Final Master Merge
master_data <- census_wide %>%
  left_join(cdc_wide,    by = "GEOID") %>%
  left_join(usda_clean,  by = "GEOID") %>%
  left_join(chrr_clean,  by = "GEOID")

# Final Verification
message("✅ Master dataset created with ", nrow(master_data), " rows.")
```

Successfully joined health, food access, and socioeconomic data into a single, analysis-ready structure

There are approximately 3,140 to 3,240 counties in the U.S. 
~3,222 rows have effectively captured the entire population of the United States at the county level

```{r}
# Save as a CSV for easy access or RDS for R-specific efficiency
write.csv(master_data, "data/processed/master_dataset.csv", row.names = FALSE)
saveRDS(master_data, "data/processed/master_dataset.rds")
```

## Install packages to R environment 

```{r}
# Install if missing
if(length(new.packages)) install.packages(new.packages)

# Load all packages at once
lapply(list.of.packages, library, character.only = TRUE)
```

## Data Limitations

Data from multiple years (2019–2025) were integrated to provide the most comprehensive county-level view. Socioeconomic metrics (ACS) represent the most recent reliable 5-year estimates (2023), while health indicators reflect the latest available public data (2025).

# References